### 2014-04-16 text/Chapter03.tex
Finish flat dataflow in chapter 3

\begin{figure}[t]
\begin{equation*}
\tyrule{var}
  {x : \tau \in \Gamma}
  {\coctx{\Gamma}{\aclrd{ 0 }} \vdash x : \tau }
\end{equation*}
\begin{equation*}
\tyrule{prev}
  {\coctx{\Gamma}{n} \vdash e : \tau}
  {\coctx{\Gamma}{n+1} \vdash \kvd{prev}~e : \tau}
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{n'}} \vdash e : \tau }
  {\coctx{\Gamma}{\aclrd{n}} \vdash e : \tau }\quad\quad(\aclrd{n'} \leq \aclrd{n})
\end{equation*}
\begin{equation*}
\tyrule{app}
  {\coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{p}} \tau_2 &
   \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }
  {\coctx{\Gamma}{\textnormal{max}(\aclrd{m}, \aclrd{n} + \aclrd{p})} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma}{\aclrd{n}+\aclrd{m}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 }
\end{equation*}

@@

\paragraph{Example.} 
In the following example, \ident{counter} generates positive integers 
and \ident{tick} flips between $0$ and $1$:
%
\begin{equation*}
\begin{array}{l}
(\,   \kvd{if}~~(\kvd{prev}~\ident{tick})=0~\kvd{then}~(\lambda x \rightarrow \kvd{prev}~x)\\
\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\quad(\kvd{prev}~\ident{counter})
\end{array}
\end{equation*}
%
The function on the left-hand side returns the previous value of the argument every other 
step. Applied to a stream $\langle{\ldots,4,3,2,1}\rangle$ (where $1$ is the ``current''
value) it yields the stream $\langle{\ldots,4,4,2,2}\rangle$. To obtain the function, 
we need one past value from the context (for $\kvd{prev}~\ident{tick}$). The returned function 
needs either none or one past value (thus a subtyping rule is required to type it
as requiring one past value).

The function is called with $\kvd{prev}~\ident{counter}$ as an argument, meaning that the result
is either the first or second past element. Given 
$\ident{counter}\hspace{-0.1em}=\hspace{-0.1em}\langle{\ldots,5,4,3,2,1}\rangle$, the argument 
is $\langle{\ldots,5,4,3,2}\rangle$ and so the result is a stream $\langle{\ldots,5,5,3,3}\rangle$.

How many past values does the whole expression need? The argument requires one past value and
the function one additional ($\kvd{prev}~(\kvd{prev}~\ident{counter}))$, adding to $2$. To obtain
the function, one past value is needed. Overall we need the maximum of what is needed to evaluate the 
function and what is needed by passing the argument to the function: $max(1, 2)=2$. This is 
captured by (\emph{app}) in Figure~\ref{figure:intro-dataflow}.

The (\emph{fun}) rule again does not allow delaying of context requirements. If the body 
requires $m$ past values, these have to be available both from the outer context and from the
argument provided by the caller. Later, we make the rule more flexible to make it compatible
with (\emph{fun}) for implicit parameters that \emph{splits} requirements.

\paragraph{Semantics.}
In this setting, a function $\tau_1 \rightarrow
\tau_2$ becomes a function from a list of historical values $[\tau_1]
\rightarrow \tau_2$. A coeffect system here tracks how many past values
to cache.

>>>>

\begin{figure}[t]
\begin{equation*}
\tyrule{var}
  {x : \tau \in \Gamma}
  {\coctx{\Gamma}{\aclrd{ 0 }} \vdash x : \tau }
\end{equation*}
\begin{equation*}
\tyrule{prev}
  {\coctx{\Gamma}{ \aclrd{n} } \vdash e : \tau}
  {\coctx{\Gamma}{ \aclrd{n}+1} \vdash \kvd{prev}~e : \tau}
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{n'}} \vdash e : \tau }
  {\coctx{\Gamma}{\aclrd{n}} \vdash e : \tau }\quad\quad(\aclrd{n'} \leq \aclrd{n})
\end{equation*}
\begin{equation*}
\tyrule{app}
  {\coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{p}} \tau_2 &
   \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }
  {\coctx{\Gamma}{\textnormal{max}(\aclrd{m}, \aclrd{n} + \aclrd{p})} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma}{\aclrd{n}+\aclrd{m}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 }
\end{equation*}

@@

\paragraph{Example.} 
As with the liveness example, the application rule might require more explanation. The following 
example is somewhat arbitrary, but it demonstrates the rule well. We assume that \ident{counter} 
is a stream of positive integers (starting from zero) and \ident{tick} flips between $0$ and $1$.
The full typing derivation is shown in Appendix~?:
%
\begin{equation*}
\begin{array}{l}
(\,   \kvd{if}~~(\kvd{prev}~\ident{tick})=0\\[-0.25em]
\,\,\, \kvd{then}~(\lambda x \rightarrow \kvd{prev}~x)\\[-0.25em]
\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\qquad(\kvd{prev}~\ident{counter})
\end{array}
\end{equation*}
%
The left-hand side of the application returns a function depending on the \emph{previous}
value of \ident{tick}. The resulting stream of functions flips between a function returning
a current value and a function returning the previous value. If the current \ident{tick} is 0, and
the function is applied to a stream $\langle{\ldots,4,3,2,1}\rangle$ 
(where $1$ is the current value), it yields the stream $\langle{\ldots,4,4,2,2}\rangle$.

To obtain the function, we need one past value from the context (for $\kvd{prev}~\ident{tick}$). The 
returned function needs either none or one past value (thus a subtyping rule is required to type 
it as requiring one past value). So, the annotations for (\emph{app}) are $\aclrd{m}=1, \aclrd{p}=1$.
The function is called with $\kvd{prev}~\ident{counter}$ as an argument, meaning that the result
is either the first or second past element. Given 
$\ident{counter}\hspace{-0.1em}=\hspace{-0.1em}\langle{\ldots,5,4,3,2,1}\rangle$, the argument 
is $\langle{\ldots,5,4,3,2}\rangle$ and so the overall result is a stream $\langle{\ldots,5,5,3,3}\rangle$.
From the argument, we get the requirement $\aclrd{n}=1$.

Using the (\emph{app}) rule, we get that the overall number of past elements needed is
$\mathit{max}(1, 1+1) = 2$. This should match the intuition about the code -- when the first function
is applied to the argument, the computation will first access $\kvd{prev}~\ident{tick}$ (using one
past value) and then $\kvd{prev}~(\kvd{prev}~\ident{counter}))$ (using two past values).

\begin{figure}
\begin{equation}
\tag{\emph{var}}
\hspace{-1em}
\sem{\coctx{\Gamma}{\aclrd{0}} \vdash x_i : \tau_i} = \lambda \langle(x_0, \ldots, x_n)\rangle \rightarrow x_i
\end{equation}
\vspace{-1.0em}
\begin{equation}
\tag{\emph{prev}}
\hspace{-1.5em}
\begin{array}{l}
\sem{\coctx{\Gamma}{\aclrd{n}+1} \vdash \kvd{prev}~e : \tau} = \lambda \langle \mathbf{v}_0, \ldots, \mathbf{v}_{\aclrd{n}+1} \rangle \rightarrow\\[-0.1em]
  \qquad\sem{ \coctx{\Gamma}{ \aclrd{n} } \vdash e : \tau }~\langle \mathbf{v}_1, \ldots, \mathbf{v}_{ \aclrd{n}+1 }\rangle
\end{array}
\end{equation}
\vspace{-0.5em}
\begin{equation}
\tag{\emph{sub}}
\hspace{-1.5em}
\begin{array}{l}
\sem{\coctx{\Gamma}{\aclrd{n}} \vdash e : \tau} = \lambda \langle \mathbf{v}_0, \ldots, \mathbf{v}_{\aclrd{n}} \rangle \rightarrow\\[-0.1em]
  \qquad\sem{ \coctx{\Gamma}{ \aclrd{n'} } \vdash e : \tau }~\langle \mathbf{v}_0, \ldots, \mathbf{v}_{ \aclrd{n'} }\rangle
\end{array}
\end{equation}
\vspace{-0.5em}
\begin{equation}
\tag{\emph{abs}}
\hspace{-1.5em}
\begin{array}{l}
  \sem{\coctx{\Gamma}{\aclrd{n}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 } = 
    \lambda \langle \mathbf{v}_0, \ldots \mathbf{v}_{\aclrd{n}} \rangle \rightarrow\\[-0.1em]
  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{n}} \vdash e : \tau_2 
    }~\langle (\mathbf{v}_0, y_0), \ldots, (\mathbf{v}_{ \aclrd{n}  }, y_{ \aclrd{n} } ) \rangle
\end{array}
\end{equation}
\vspace{-0.5em}
\begin{equation*}
\tag{\emph{app}}
\hspace{-1.5em}
\begin{array}{l}
  \sem{\coctx{\Gamma}{\textnormal{max}(\aclrd{m}, \aclrd{n}+\aclrd{p})} \vdash e_1~e_2 : \tau_2 } = 
    \lambda (\mathbf{v}_0, \ldots, \mathbf{v}_{ \textnormal{max}(\aclrd{m}, \aclrd{n}+\aclrd{p}) } ) \rightarrow \\[-0.1em]
  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{p}} \tau_2 
    }~(\mathbf{v}_0, \ldots, \mathbf{v}_{ \aclrd{m} })\\[-0.1em]
  \qquad\kvd{in}~g~(~\sem{ \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }~(\mathbf{v}_0, \ldots, \mathbf{v}_{ \aclrd{n} }),\ldots, \\[-0.1em]
  \qquad\qquad\;~\sem{ \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }~(\mathbf{v}_{ \aclrd{p} }, \ldots, \mathbf{v}_{ \aclrd{n}+\aclrd{p} })~)
\end{array}
\end{equation*}

\caption{Semantics of a language with implicit parameters}
\label{fig:applications-flat-dfsem}
\end{figure}

\paragraph{Semantics.}
The sample language discussed in this section is a \emph{causal} data-flow language. This means that
a computation can access \emph{past} values of the stream (but not future values). In the semantics, 
we again need richer structure over the input.

Uustalu and Vene \cite{comonads-notions} model causal data-flow computations using a non-empty list
$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$ over the input. A function $\tau_1 \rightarrow \tau_2$
is thus modelled as $\ident{NeList}~\tau_1 \rightarrow \tau_2$. This model is difficult to implement
efficiently, as it creates unbounded lists of past elements.

The coeffect system tracks maximal number of past values and so we can define the semantics using
a list of fixed length. As with liveness, this is a data structure \emph{indexed} by the coeffect
annotation. We write $\tau^{\aclrd{n}}$ for a list containing $\aclrd{n}$ elements (which can be 
also viewed as an $\aclrd{n}$-element product $\tau \times \ldots \times \tau$).

As with the previous examples, our semantics interprets a judgement using a (semantic) function;
functions in the language are modelled as functions taking a list of inputs:
%
\begin{equation*}
\begin{array}{rcl}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \aclrd{n} } \vdash e : \tau} 
  &:& (\tau_1 \times \ldots \times \tau_n)^{\aclrd{n}+1} \rightarrow \tau\\
\sem{\tau_1 \xrightarrow{\aclrd{n}} \tau_2} &:& \tau_1^{\aclrd{n}+1} \rightarrow \tau_2
\end{array}
\end{equation*}
%
Note that the semantics requires one more value than is the number of past values. This is because 
the first value is the current value and has to be always available, even when the annotation is
zero as in (\emph{var}).

The rules defining the semantics are shown in Figure~\ref{fig:applications-flat-dfsem}. The
semantics of the context is a \emph{list of pairs}. To make the rules easier to follow, we write
$\langle \mathbf{v}_1, \ldots, \mathbf{v}_n \rangle$ for an $n$-element list containing pairs.
Pairs that model the entire context such as $\mathbf{v}_1$ are written in bold. When we access
individual variables, we write $\mathbf{v} = (x_1, \ldots, x_m)$ where $x_i$ denote individual
variables of the context.

In (\emph{var}), the context is a singleton-list containing a product of variables, from which
we project the right one. In (\emph{prev}) and (\emph{sub}), we drop some of the elements from 
the history (from the front and end, respectively) and then evaluate the original expression.

Lambda abstractions (\emph{abs}) receives two lists of the same size -- one containing values of
the variables from the declaration-site $\langle \mathbf{v}_0, \ldots, \mathbf{v}_{\aclrd{n}} \rangle$ and one
containing the argument provided by the call-site $\langle y_0, \ldots, v_{\aclrd{n}} \rangle$.
The semantics applies the well-known \emph{zip} operation on the lists and passes the result to the
body.

Finally, application (\emph{abs}) uses the input context in two ways, which gives rise to the
two requirements combined using \emph{max}. First, it evaluates the expression $e_1$ which is 
called with the past $\aclrd{m}$ values. The resulting function $g$ is then sequentially composed
with the semantics of $e_2$. To call the function, we need to evaluate $e_2$ repeatedly -- namely,
$\aclrd{p}+1$ times, which results in the overall requirement for $\aclrd{n}+\aclrd{p}$ past values.

\paragraph{Summary}

\todo{Key points...}
